Synthesising Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-07-09 papers/termination/paper.tex
Remove author names for double blind
post-submission typo spotting

\authorinfo{Cristina David\and Daniel Kroening\and Matt Lewis}
           {University of Oxford}
           {firstname.lastname@cs.ox.ac.uk}

@@

\begin{itemize}
 \item Our transition relation is $T(x, x') \subseteq X \times X$.
 \item Our loop condition is $L(x) \subseteq X$.
 \item Our ranking function is $R(x) : X \to Y$.
 \item Our state space has size $\| X \| = M = 2^k$.
 \item Our ranking co-domain has size $\| Y \| = N = 2^j$.
 \item The number of looping states is $\| L \| = l$.
 \item Our transition relation is deterministic and parititioned into chains of length $c_i$, with $l = \sum c_i$.
\end{itemize}

@@

This extra second-order term introduces some complexity to the formula, which
we can avoid if the transition relation $T$ is deterministic.
\begin{definition}[Determinism]
 A relation $T$ is deterministic iff each state $x$ has exactly one successor under $T$:
 $$\forall x . \exists x' . T(x, x') \wedge \forall x'' . T(x, x'') \rightarrow x'' = x'$$
\end{definition}
In order to describe a deterministic \emph{program} in a way that still allows us
to sensibly talk about termination, we assume the existence of a special sink
state $s$ with no outgoing transitions and such that $\lnot G(s)$ for any
of the loop guards $G$.  The program is deterministic if its transition
relation is determinstic for all states except $s$.

@@

\begin{theorem}
\label{thm:ct}
 The loop $L(G, T)$ terminates when started in any state satisfying $I(x)$ iff formula {\bf [CT]}
 (Defintion~\ref{def:ct}, Figure~\ref{fig:conditional_termination}) is satisfiable.
\end{theorem}
% 
% \begin{proof}
%  Do this proof.
% \end{proof}

@@

%\begin{figure}
\begin{algorithm}
{\small
 \caption{Abstract synthesis algorithm
 \label{alg:cegis}}
 %\vspace{-1.7em}
 %\begin{multicols}{2}
 \begin{algorithmic}[1]
 \Statex
\Function{synth}{{\sc inputs}, $\sigma$}
  \Let{$(i_1, \ldots, i_N)$}{{\sc inputs}}
  \Let{query}{$\exists P, x_0 . \forall i \in \text{\sc inputs} . \sigma(P, x_0, i)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{unsatisfiable}
  \EndIf
\EndFunction
\Statex
\Function{verif}{$P, x_0, \sigma$}
  \Let{query}{$\exists x . \lnot \sigma(P, x_0, x)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{valid}
  \EndIf
\EndFunction
\columnbreak
\Statex
\Function{Synthesise}{$\sigma$}
  \Let{{\sc inputs}}{$\emptyset$}
  \Loop
    \Let{candidate}{\Call{synth}{{\sc inputs}}}
    \If{candidate = UNSAT}
      \State \Return{unsatisfiable}
    \EndIf
    \Let{res}{\Call{verif}{candidate}}
    \If{res = valid}
      \State \Return{candidate}
    \Else
      \Let{{\sc inputs}}{{\sc inputs} $\cup$ res}
    \EndIf
  \EndLoop
\EndFunction
 \end{algorithmic}
 %\end{multicols}
 }
\end{algorithm}
%\end{figure}

@@

We will now show that the number of iterations of the CEGIS loop is a function
of the Kolmogorov complexity of the \mbox{(non-)}ter\-mi\-nation proof, and so
our runtime is asymptotically a function of the Kolmogorov complexity.
We argue that this gives our procedure various desirable qualities in practical
appliations.
We first recall the definition of the Kolmogorov complexity
of a function $f$:

@@

\nocite{fisher2009phylogeny}

>>>>

%\authorinfo{Cristina David\and Daniel Kroening\and Matt Lewis}
%           {University of Oxford}
%           {firstname.lastname@cs.ox.ac.uk}

%\authorinfo{}

@@

\begin{itemize}
 \item Our transition relation is $T(x, x') \subseteq X \times X$.
 \item Our loop condition is $L(x) \subseteq X$.
 \item Our ranking function is $R(x) : X \to Y$.
 \item Our state space has size $\| X \| = M = 2^k$.
 \item Our ranking co-domain has size $\| Y \| = N = 2^j$.
 \item The number of looping states is $\| L \| = l$.
 \item Our transition relation is deterministic and partitioned into chains of length $c_i$, with $l = \sum c_i$.
\end{itemize}

@@

This extra second-order term introduces some complexity to the formula, which
we can avoid if the transition relation $T$ is deterministic.
\begin{definition}[Determinism]
 A relation $T$ is deterministic iff each state $x$ has exactly one successor under $T$:
 $$\forall x . \exists x' . T(x, x') \wedge \forall x'' . T(x, x'') \rightarrow x'' = x'$$
\end{definition}
In order to describe a deterministic \emph{program} in a way that still allows us
to sensibly talk about termination, we assume the existence of a special sink
state $s$ with no outgoing transitions and such that $\lnot G(s)$ for any
of the loop guards $G$.  The program is deterministic if its transition
relation is deterministic for all states except $s$.

@@

\begin{theorem}
\label{thm:ct}
 The loop $L(G, T)$ terminates when started in any state satisfying $I(x)$ iff formula {\bf [CT]}
 (Definition~\ref{def:ct}, Figure~\ref{fig:conditional_termination}) is satisfiable.
\end{theorem}
% 
% \begin{proof}
%  Do this proof.
% \end{proof}

@@

%\begin{figure}
\begin{algorithm}
{\small
 \caption{Abstract synthesis algorithm
 \label{alg:cegis}}
 %\vspace{-1.7em}
 %\begin{multicols}{2}
 \begin{algorithmic}[1]
 \Statex
\Function{synth}{{\sc inputs}, $\sigma$}
  \Let{$(i_1, \ldots, i_N)$}{{\sc inputs}}
  \Let{query}{$\exists P, x_0 . \forall i \in \text{\sc inputs} . \sigma(P, x_0, i)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{unsatisfiable}
  \EndIf
\EndFunction
\Statex
\Function{verif}{$P, x_0, \sigma$}
  \Let{query}{$\exists x . \lnot \sigma(P, x_0, x)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{valid}
  \EndIf
\EndFunction
%\columnbreak
\Statex
\Function{Synthesise}{$\sigma$}
  \Let{{\sc inputs}}{$\emptyset$}
  \Loop
    \Let{candidate}{\Call{synth}{{\sc inputs}}}
    \If{candidate = UNSAT}
      \State \Return{unsatisfiable}
    \EndIf
    \Let{res}{\Call{verif}{candidate}}
    \If{res = valid}
      \State \Return{candidate}
    \Else
      \Let{{\sc inputs}}{{\sc inputs} $\cup$ res}
    \EndIf
  \EndLoop
\EndFunction
 \end{algorithmic}
 %\end{multicols}
 }
\end{algorithm}
%\end{figure}

@@

We will now show that the number of iterations of the CEGIS loop is a function
of the Kolmogorov complexity of the \mbox{(non-)}ter\-mi\-nation proof, and so
our runtime is asymptotically a function of the Kolmogorov complexity.
We argue that this gives our procedure various desirable qualities in practical
applications.
We first recall the definition of the Kolmogorov complexity
of a function $f$:

@@

%\nocite{fisher2009phylogeny}
